Methods

### 2019-10-21 kellen/Legends.Rmd
current edits

**Figure 1. Performance is mildly dependent on layers**. A feed-forward neural network with either one or two hidden layer(s) and 16 nodes was used to predict TAD boundaries. Adjusted PPV (200kb flank) was calculated to determine the accuracy of predictions. Results are seperated by number of layers (1 or 2) and predictor type (Combined, Distance, Overlap, Percent Overlap). GM12878 cell line, various resolutions.

**Figure 2. Performance is mildly dependent on dropout**. A feed-forward neural network with one layer and 16 nodes was used to predict TAD boundaries. Adjusted PPV (200kb flank) was calculated to determine the accuracy of predictions. Varying levels of dropout were used (0,0.1,0.2,0.3,0.4,0.5) Results are seperated by dropout level and predictor type (Combined, Distance, Overlap, Percent Overlap). GM12878 cell line, various resolutions.

**Figure 3. Over or under sampling is needed to account for class imbalance**. A feed-forward neural network with a single hidden layer and 16 nodes was used to predict TAD boundaries. Adjusted PPV was calculated to determine the of accuracy predictions. Results are seperated by oversampling approach (None, RUS, ROS, SMOTE) and feature type (TFBS, BroadHMM, Histones). All results were generated using distance. GM12878 cell line, various resolutions.

**Figure 4. Results are generally consistent with node selection**. A feed-forward neural network with a single hidden layer and varying numbers of nodes (4,8,12,16,24,32,64) was used to predict TAD boundaries. Adjusted PPV was calculated to determine the accuracy of predictions. Results are presented for TFBS (a), histones (b) and broadHMM (c) and feature type (Combined, Distance, Overlap and Percent Overlap). GM12878 cell line, various resolutions.

>>>>

**Figure 1. Performance is mildly dependent on layers**. A feed-forward neural network with either one or two hidden layer(s) and 16 nodes was used to predict TAD boundaries. Adjusted PPV (200kb flank) was calculated to determine the accuracy of predictions. Results are seperated by number of layers (1 or 2) and predictor type (Combined, Distance, Overlap, Percent Overlap). GM12878 cell line, various resolutions. [??? No dropout layers?]

**Figure 2. Performance is mildly dependent on dropout**. A feed-forward neural network with one layer and 16 nodes was used to predict TAD boundaries. Adjusted PPV (200kb flank) was calculated to determine the accuracy of predictions. Varying levels of dropout were used (0, 0.1, 0.2, 0.3, 0.4, 0.5) Results are separated by dropout level and predictor type (Combined, Distance, Overlap, Percent Overlap). GM12878 cell line, various resolutions.

**Figure 3. Over or under sampling is needed to account for class imbalance**. A feed-forward neural network with a single hidden layer and 16 nodes [??? Dropout?] was used to predict TAD boundaries. Adjusted PPV was calculated to determine the accuracy of predictions. Results are separated by oversampling approach (None, RUS, ROS, SMOTE) and feature type (TFBS, BroadHMM, Histones). All results were generated using distance. GM12878 cell line, various resolutions.

**Figure 4. Results are generally consistent with node selection**. A feed-forward neural network with a single hidden layer and varying numbers of nodes (4, 8, 12, 16, 24, 32, 64) was used to predict TAD boundaries. Adjusted PPV was calculated to determine the accuracy of predictions. Results are presented for TFBS (a), histones (b) and broadHMM (c) and feature type (Combined, Distance, Overlap and Percent Overlap). GM12878 cell line, various resolutions.

### 2019-10-21 kellen/manuscript.Rmd
current edits

## Choosing the neural network hyperparameters

A feed forward neural network was fit to predict the location of TAD boundaries. The ideal number of hidden layers (1, 2) and nodes (4, 8, 12, 16, 32, 64), and the resampling scheme ("none", "SMOTE", "RUS", "ROS") was tested based on cross-validation [??? Need details, how cross-validation was set up] using adjusted precision as our measure of accuracy. Results were summarized across all chromosomes. To prevent overfitting, a dropout layer was added between every layer. In general, we find that networks with more than 1 layer tend to either overfit the data (Figure 1), resulting in a decrease in prediction quality. As a result, we present results for single layer models exclusively. Additionally, we find that a dropout of 50% on the final layer provides consistently better results (Figure 2). In addition, we find the choice of activation function to be largely inconsequential to our predictive ability as a result a simple ReLU activation function is used for all layers.

@@

TAD boundary data has two unique features, it is heavily imbalanced (there are much less TAD boundaries than non-TAD boundaries) and it is imprecise due to limitations with resolution. An additional complication arises due to the fact that TAD callers are only capable of identifying single points as their best guess for a TAD boundary. As a result, the response vector of any predictive algorithm will be comparitively sparse compared to the features.

The implicit impreciseness of TAD calling leads to inflated false positive values as algorithms will detect TADs not only at the detected TAD boundary but also in regions around it. As a result, traditional measures of prediction quality will return poor results for any algorithm. For this reason, we introduce the flanking region around the boundary allowing us to account for boundaries predicted near TAD boundaries but not exactly on the boundary. Precision is used because it monotonically increases with flank size due to the absence of false negative and true negatives in its formula. This is not the case for measurements such as accuracy, F1 and matthews correlation coefficient which can decrease with increasing flank size due to the removal of negative cases as the flank becomes larger.

@@

We find that SMOTE oversampling slightly outperforms other oversampling methods in terms of TAD boundary detection (Figure 3, Supplementary Table 1). Summarized across all resolutions, feature types and data sources, SMOTE performed best (Mean $PPV_a$ = 0.837, Median $PPV_a$ = 0.933) with a slight advantage over RUS (Mean $PPV_a$ = 0.837, Median $PPV_a$ = 0.930). In general, we find that the all oversampling and undersampling methods produce comparable results. However, we find that any oversampling method results in better precision than no oversampling at all.

Another side-effect of imbalance is algorithms failing to learn the data and simply predicting no positive outcomes (TAD boundaries). To test, how well oversampling approaches deal with this issue we fit models across all resolutions (5kb,10kb,25kb,50kb,100kb), feature types (distance, overlap, percent overlap, combined) and data sources (TFBS, histones, broadHMM) and calculated the percentage of models with zero boundary predictions. In total, there were 1320 models tested for each sampling type. Without oversampling or undersampling, the vast majority of models (74%) fail to predict a single TAD boundary across all data types and resolutions. SMOTE has the smallest proportion of failures (0.5%) followed by RUS (0.9%) and ROS (2%). This demonstrates the need to balance the data before predicting TAD boundaries.

[??? figure 1 has results for "none" only for 5Kb resolution - why?? The rest are zero or below the threshold]
[??? Adjusted PPV are too similar. See above, we need a strong justification why not AUC, MCC, F1. Can you try if any of those can be used? I am quite uneassy that we'll be asked to use something else than PPV. So, justify]
[??? Figure 1 - can we quantify that on average SMOTE is better? It doesn't look so?? Added but very slight]
[??? Figure 1 - can we quantify TFBSs are better than others?? Added later but very slight]

## Network architecture engineering

We find that regardless of the number of hidden layers (4,8,12,16,32,64), the neural network is able to predict TAD boundaries near those annotated by TAD callers. The number of layers with the best performance is 16 (Median $PPV_a$ = .941) and the worst is 4 (Median $PPV_a$ = .932). In rare circumstances (Figure 4a,4b,4c, Supplementary Table 2), we find that networks with too few nodes will have poor predictive ability. As a result, we use a single layer network with 16 layers as our default architecture for comparison of features. Using results from testing layers (Figure 1) and dropout (Figure 2), we define a baseline model for testing with a single hidden layer containg 16 nodes and a final dropout of 0.5. This is used for all analysis unless otherwise stated.

>>>>

## Choosing the neural network hyperparameters [??? This should be a part of network architecture engineering]

A feed forward neural network was fit to predict the location of TAD boundaries. The ideal number of hidden layers (1, 2) and nodes (4, 8, 12, 16, 32, 64), and the resampling scheme ("none", "SMOTE", "RUS", "ROS") was tested based on cross-validation [??? Need details, how cross-validation was set up] using adjusted precision as our measure of accuracy. Results were summarized across all chromosomes. To prevent overfitting, a dropout layer was added between every layer. In general, we find that networks with more than 1 layer tend to either overfit the data (Figure 1) [??? As judged by the slight decrease of PPV?], resulting in a decrease in prediction quality. As a result, we present results for single layer models exclusively. Additionally, we find that a dropout of 50% on the final layer provides consistently better results (Figure 2). [??? Extremely hard to see. Can it be quantified?] In addition, we find the choice of activation function [??? Opens Pandora's box - need to show the proof. What other hunctions did you compare] to be largely inconsequential to our predictive ability as a result a simple ReLU activation function is used for all layers.

@@

TAD boundary data has two [??? Unclear, what are the two features?] unique properties affecting prediction performance. The first property is that the data is heavily imbalanced (there are much less TAD boundaries than non-TAD boundaries). and it is imprecise due to limitations with resolution. An additional complication arises due to the fact that TAD callers are only capable of identifying single points as their best guess for a TAD boundary [??? I don't understand]. As a result, the response vector of any predictive algorithm will be comparitively sparse compared to the features [??? Unclear, which features].

The implicit impreciseness of TAD calling leads to inflated false positive values as algorithms will detect TADs not only at the detected TAD boundary but also in regions around it. As a result, traditional measures of prediction quality [??? like what?] will return poor results for any algorithm. For this reason, we introduce the flanking region around the boundary [??? It is best to be illustrated. I don't get this idea how it will affect performance] allowing us to account for boundaries predicted near TAD boundaries but not exactly on the boundary. Precision is used because it monotonically increases with flank size due to the absence of false negative and true negatives in its formula. [??? I do not understand, too many different statements] This is not the case for measurements such as accuracy, F1 and matthews correlation coefficient which can decrease with increasing flank size due to the removal of negative cases as the flank becomes larger.

@@

We find that SMOTE oversampling slightly outperforms other oversampling methods in terms of TAD boundary detection (Figure 3, Supplementary Table 1). Summarized across all resolutions, feature types and data sources, SMOTE performed best (Mean $PPV_a$ = 0.837, Median $PPV_a$ = 0.933) with a slight advantage over RUS (Mean $PPV_a$ = 0.837, Median $PPV_a$ = 0.930). In general, we find that the all oversampling and undersampling methods produce comparable results and increase prediction performance as compared with no oversampling.

Another side-effect of imbalance is algorithms failing to learn the data and simply predicting no positive outcomes (TAD boundaries). To test, how well oversampling approaches deal with this issue we fit models across all resolutions (5kb, 10kb, 25kb, 50kb, 100kb), feature types (distance, overlap, percent overlap, combined) and data sources (TFBS, histones, broadHMM [??? We don't combine them? TBD, but we should]) and calculated the percentage of models with zero boundary predictions. In total, there were 1,320 [??? Need details how this number was obtained. It does not account] models tested for each sampling type. Without oversampling or undersampling, the vast majority of models (74%) fail to predict a single TAD boundary across all data types and resolutions. SMOTE has the smallest proportion of failures (0.5%) followed by RUS (0.9%) and ROS (2%). This demonstrates the need to balance the data before predicting TAD boundaries.

## Network architecture engineering [??? Combine it with the last Methods section]

We find that regardless of the number of hidden layers (4, 8, 12, 16, 32, 64), the neural network is able to predict TAD boundaries near those annotated by TAD callers. The number of layers with the best performance is 16 (Median $PPV_a$ = .941) and the worst is 4 (Median $PPV_a$ = .932). In rare circumstances (Figure 4a, 4b, 4c, Supplementary Table 2), we find that networks with too few nodes will have poor predictive ability. As a result, we use a single layer network with 16 layers as our default architecture for comparison of features. Using results from testing layers (Figure 1) and dropout (Figure 2), we define a baseline model for testing with a single hidden layer containg 16 nodes and a final dropout of 0.5. This is used for all analysis unless otherwise stated.
